---
title: "PH125.9x Data Science Capstone Final Movie Project Report"
author: "Marina Ganopolsky"
date: "8/6/2020"
output: 
  pdf_document:
    latex_engine: xelatex
mainfont: Montserrat
toc: true
number_sections: true
toc_depth: 1
df_print: kable
fontsize: 13pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Executive Summary

This is Marina Ganopolsky's implementation of the **MovieLens Project** of the **HervardX: PH125.9x Data Science, Summer 2020: Capstone**.

**The objectives of this project are:**
* To create a movie recommendation engine based on the data set provided by training a set of algorithm on provided wrangled training data
* The engine will choose the best-performing algorithm, and make a final evaluation on the validation dataset 
* Test the recommendations with the RMSE formula
  + Find an **RMSE**  (Root Mean Square Error) of the predicted ratings, **ideally < 0.86490**, and validate this with the validation data set, as described below.

The dataset is provided by the staff, and is wrangled by me for better use during the assignment beyond the manipulations suggested in the assignment; the modifications will be shown in the **Data** section below. 

A quick summary of the data wrangling modifications will be provided, as well as some visually representations of patterns present in the data in the **Data** section as well. This will inform the reader about the movies, users and general trends in the dataset. 

An exploratory analysis will be performed to generate predicted movie ratings, with various models, to be specified, until a final RMSE calculation fits our requirements. Results of the RMSE calculations will be analyzed and explained, and a conclusion will be provided.

## Aim of the project

The aim in this project is to train a machine learning algorithm that predicts user ratings (from 0.5 to 5 stars) using the inputs of a provided subset  (edx dataset provided by the staff) to predict movie ratings in a provided validation set.

The value used to evaluate algorithm performance is the Root Mean Square Error, or RMSE. RMSE is one of the most used measure of the differences between values predicted by a model and the values observed. RMSE is a measure of accuracy, to compare forecasting errors of different models for a particular dataset, a lower RMSE is better than a higher one. The effect of each error on RMSE is proportional to the size of the squared error; thus larger errors have a disproportionately large effect on RMSE. Consequently, RMSE is sensitive to outliers.
Four models that will be developed will be compared using their resulting RMSE in order to assess their quality. The evaluation criteria for this algorithm is a RMSE expected to be lower than 0.8775.
The function that computes the RMSE for vectors of ratings and their corresponding predictors will be the following:
$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$

```{r RMSE_function1, echo = FALSE}
RMSE <- function(predicted_ratings, true_ratings){
  sqrt(mean((predicted_ratings - true_ratings)^2))
}
```


Finally, the best resulting model will be used to predict the movie ratings.


## Dataset

The MovieLens dataset is automatically downloaded with the code provided by the staff; in the interests of brevity I will not be including the code here. The datasets are also provided, and can be additionaly downloaded here:

• [MovieLens 10M dataset]  https://grouplens.org/datasets/movielens/10m/

• [MovieLens 10M dataset - zip file] http://files.grouplens.org/datasets/movielens/ml-10m.zip

As specified by the project description (and the provided code) the data is split into a validation set and the "rest". I further split the data, after it was manipulated, into a testing and training sets; The calculations and algorithm verifications are done on the testing and training sets, and after the final RMSE is chosen, this algorithm is applied to the validation test to verify the hypothesis.

A fair amount of data manipulation has been done to the datasets created with the provided code. The changes include:
* Adding a ratings average per movie in **avg_rating**
* Extracting the release year of the movie from its title. (Every movie title seems to include the release year in at the end of title, surrounded by parentheses.)
  + The field **release_year** is created with a ReGex ```{r eval=FALSE} "(\\(\\d{4}\\)$)" ```
* Removing the year from the movie title
* Removing the parentheses from the **release_year** and converting it to a numerical field
* Creating a field called **rating_date** , by extracting the rating date from the **timestamp** field
* Creating a field called **age_of_movie**, by subtracting the year the film was released from the year parameter of the current date, acquired by ```{r eval=FALSE} year(as_date(Sys.Date()))```
* Creating a field called **rating_year** by extracting the rating year from the **rating_date** field.
* As provided, the film genres are provided as a string separated by "|". Therefor, I further manipulate the movielens dataset (and significantly increase its size) by extracting out the film genres, with the following code:
```{r eval=FALSE} movielens <- movielens %>% separate_rows(genres, sep ="\\|")```
* After the data wrangling is complete, the dataset is split up into :
  + **A validation set**, comprised of **10%** of the data complete data
    + this portion of the data shall not to be analyzed until the final RMSE is found to be within acceptable range 
  + **A testing dataset**
  + **A training dataset**

In order to predict in the most accurate way the movie rating of the users that haven’t seen the movie yet, the he MovieLens dataset will be splitted into 2 subsets that will be the “edx”, a training subset to train the algorithm, and “validation” a subset to test the movie ratings.  

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(stringr)
library(lubridate)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
marinas_directory = "/Users/marina/Documents/DevProjects/R/projects/r_final_harvard_data_science_project"

#specifying not needing to download the .zip file if located on personal computer
if (file.exists(marinas_directory)){
  setwd(marinas_directory)
}


if (file.exists("ml-10m.zip")) {
  
  file.link("./ml-10m.zip", dl)
} else {
  #only download the file if it doesn't exist in the current working directory.
  download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
}

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later
#mutate to add the "year" column, and remove the (year) pattern from the title 

movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres)) %>%
  mutate(release_year =  str_extract(title, "(\\(\\d{4}\\)$)") ) %>%
  mutate(release_year = as.numeric(str_remove(str_remove(release_year, "[\\(]"), "[\\)]")) ,
         title = str_trim(str_remove(title, "\\(\\d{4}\\)$")))

# add the date field , converting the timestamp to an actual date, and rounding to the nearest day
#use inner_join instead of provided left_join to make sure the movie titles and genres are populated
movielens <- inner_join(ratings, movies, by = "movieId") %>% 
  mutate(rating_date = round_date(as_datetime(timestamp), unit="day"),
         age_of_movie =  year(as_date(Sys.Date())) - release_year,
         rating_year = year(rating_date))


#separate the genres from the combined values into separate ones
movielens <- movielens %>% separate_rows(genres, sep ="\\|")

```

Algorithm development is to be carried out on the "edx" subset only, as "validation" subset will be used to test the final algorithm.
\pagebreak


# Methods and Analysis


## Data Analysis

To get familiar with the dataset, we find the first rows of "edx" subset as below.
The subset contain the six variables “userID”, “movieID”, “rating”, “timestamp”, “title”, and “genres”. Each row represent a single rating of a user for a single movie.

```{r head, echo = FALSE}
head(edx) %>%
  print.data.frame()
  
```

A summary of the subset confirms that there are no missing values.

```{r summary, echo = FALSE}
summary(edx)
```

The total of unique movies and users in the edx subset is about 70.000 unique users and about 10.700 different movies:

```{r, echo = FALSE}
edx %>%
summarize(n_users = n_distinct(userId), 
          n_movies = n_distinct(movieId))
```

Users have a preference to rate movies rather higher than lower as shown by the distribution of ratings below. 4 is the most common rating, followed by 3 and 5. 0.5 is the least common rating. In general, half rating are less common than whole star ratings.

```{r rating_distribution, echo = FALSE}
edx %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.25, color = "black") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  scale_y_continuous(breaks = c(seq(0, 3000000, 500000))) +
  ggtitle("Rating distribution")
  
```


We can observe that some movies have been rated moch often that other, while some have very few  ratings and sometimes only one rating. This will be important for our model as very low rating numbers might results in untrustworthy estimate for our predictions. In fact 125 movies have been rated only once. 

Thus regularisation and a penalty term will be applied to the models in this project. Regularizations are techniques used to reduce the error by fitting a function appropriately on the given training set and avoid overfitting (the production of an analysis that corresponds too closely or exactly to a particular set of data, and may therefore fail to fit additional data or predict future observations reliably). Regularization is a technique used for tuning the function by adding an additional penalty term in the error function. The additional term controls the excessively fluctuating function such that the coefficients don’t take extreme values.


```{r number_of_ratings_per_movie, echo = TRUE, fig.height=4, fig.width=5}
edx %>%
count(movieId) %>%
ggplot(aes(n)) +
geom_histogram(bins = 30, color = "black") +
scale_x_log10() +
xlab("Number of ratings") +
  ylab("Number of movies") +
ggtitle("Number of ratings per movie")
```


As 20 movies that were rated only once appear to be obscure, predictions of future ratings for them will be difficult.


```{r obscure_movies, echo = TRUE, fig.height=4, fig.width=5}
edx %>%
  group_by(movieId) %>%
  summarize(count = n()) %>%
  filter(count == 1) %>%
  left_join(edx, by = "movieId") %>%
  group_by(title) %>%
  summarize(rating = rating, n_rating = count) %>%
  slice(1:20) %>%
  knitr::kable()
  
```


We can observe that the majority of users have rated between 30 and 100 movies. So, a user penalty term need to be included later in our models.


```{r number_ratings_given_by_users, echo = TRUE, fig.height=4, fig.width=5}
edx %>%
count(userId) %>%
ggplot(aes(n)) +
geom_histogram(bins = 30, color = "black") +
scale_x_log10() +
xlab("Number of ratings") + 
ylab("Number of users") +
ggtitle("Number of ratings given by users")
```


Furthermore, users differ vastly in how critical they are with their ratings. Some users tend to give much lower star ratings and some users tend to give higher star ratings than average. The visualization below includes only users that have rated at least 100 movies.


```{r Mean_movie_ratings_given_by_users, echo = TRUE, fig.height=4, fig.width=5}
edx %>%
  group_by(userId) %>%
  filter(n() >= 100) %>%
  summarize(b_u = mean(rating)) %>%
  ggplot(aes(b_u)) +
  geom_histogram(bins = 30, color = "black") +
  xlab("Mean rating") +
  ylab("Number of users") +
  ggtitle("Mean movie ratings given by users") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  theme_light()
  
```


## Modelling Approach

We write now the loss-function, previously anticipated, that compute the RMSE, defined as follows:

$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$


with N being the number of user/movie combinations and the sum occurring over all these combinations.
The RMSE is our measure of model accuracy.
We can interpret the RMSE similarly to a standard deviation: it is the typical error we make when predicting a movie rating. If its result is larger than 1, it means that our typical error is larger than one star, which is not a good result.
The written function to compute the RMSE for vectors of ratings and their corresponding predictions is:


```{r RMSE_function2, echo = TRUE}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

The lower the better, as said previously.


### I. Average movie rating model

The first basic model predicts the same rating for all movies, so we compute the dataset’s mean rating. The expected rating of the underlying data set is between 3 and 4.
We start by building the simplest possible recommender system by predicting the same rating for all movies regardless of user who give it. A model based approach assumes the same rating for all movie with all differences explained by random variation :
$$ Y_{u, i} = \mu + \epsilon_{u, i} $$
with $\epsilon_{u,i}$ independent error sample from the same distribution centered at 0 and $\mu$ the “true” rating for all movies. This very simple model makes the assumption that all differences in movie ratings are explained by random variation alone. We know that the estimate that minimize the RMSE is the least square estimate of $Y_{u,i}$ , in this case, is the average of all ratings:
The expected rating of the underlying data set is between 3 and 4.

```{r, echo = TRUE}
mu <- mean(edx$rating)
mu
```


If we predict all unknown ratings with $\mu$ or mu, we obtain the first naive RMSE:

```{r naive_rmse, echo = TRUE}
naive_rmse <- RMSE(validation$rating, mu)
naive_rmse
```


Here, we represent results table with the first RMSE:

```{r rmse_results1, echo = TRUE}
rmse_results <- data_frame(method = "Average movie rating model", RMSE = naive_rmse)
rmse_results %>% knitr::kable()
```

This give us our baseline RMSE to compare with next modelling approaches.

In order to do better than simply predicting the average rating, we incorporate some of insights we gained during the exploratory data analysis.


### II.  Movie effect model

To improve above model we focus on the fact that, from experience, we know that some movies are just generally rated higher than others. Higher ratings are mostly linked to popular movies among users and the opposite is true for unpopular movies. We compute the estimated deviation of each movies’ mean rating from the total mean of all movies $\mu$. The resulting variable is called "b" ( as bias ) for each movie "i" $b_{i}$, that represents average ranking for movie $i$:
$$Y_{u, i} = \mu +b_{i}+ \epsilon_{u, i}$$

The histogram is left skewed, implying that more movies have negative effects


```{r Number_of_movies_with_the computed_b_i, echo = TRUE, fig.height=3, fig.width=4}
movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))
movie_avgs %>% qplot(b_i, geom ="histogram", bins = 10, data = ., color = I("black"),
ylab = "Number of movies", main = "Number of movies with the computed b_i")
```


This is called the penalty term movie effect.

Our prediction improve once we predict using this model.

```{r predicted_ratings, echo = TRUE}
predicted_ratings <- mu +  validation %>%
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)
model_1_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie effect model",  
                                     RMSE = model_1_rmse ))
rmse_results %>% knitr::kable()
```


So we have predicted movie rating based on the fact that movies are rated differently by adding the computed $b_{i}$ to $\mu$. If an individual movie is on average rated worse that the average rating of all movies $\mu$ , we predict that it will rated lower that $\mu$ by $b_{i}$, the difference of the individual movie average from the total average.

We can see an improvement but this model does not consider the individual user rating effect.


### III. Movie and user effect model

We compute the average rating for user $\mu$, for those that have rated over 100 movies, said penalty term user effect. In fact users affect the ratings positively or negatively.
```{r, echo = TRUE}
user_avgs<- edx %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  filter(n() >= 100) %>%
  summarize(b_u = mean(rating - mu - b_i))
user_avgs%>% qplot(b_u, geom ="histogram", bins = 30, data = ., color = I("black"))
```

There is substantial variability across users as well: some users are very cranky and other love every movie. This implies that further improvement to our model my be:
$$Y_{u, i} = \mu + b_{i} + b_{u} + \epsilon_{u, i}$$
where $b_{u}$ is a user-specific effect. If a cranky user (negative $b_{u}$ rates a great movie (positive $b_{i}$), the effects counter each other and we may be able to correctly predict that this user gave this great movie a 3 rather than a 5.

We compute an approximation by computing $\mu$ and $b_{i}$, and estimating  $b_{u}$, as the average of $$Y_{u, i} - \mu - b_{i}$$

```{r user_avgs, echo = TRUE}
user_avgs <- edx %>%
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
  
```

We can now construct predictors and see RMSE improves:


```{r model_2_rmse, echo = TRUE}
predicted_ratings <- validation%>%
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)
model_2_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie and user effect model",  
                                     RMSE = model_2_rmse))
rmse_results %>% knitr::kable()
```


Our rating predictions further reduced the RMSE. But we made stil mistakes on our first model (using only movies). The supposes “best “ and “worst “movie were rated by few users, in most cases just one user. These movies were mostly obscure ones. This is because with a few users, we have more uncertainty. Therefore larger estimates of $b_{i}$, negative or positive, are more likely.
Large errors can increase our RMSE. 

Until now, we computed standard error and constructed confidence intervals to account for different levels of uncertainty. However, when making predictions, we need one number, one prediction, not an interval. For this we introduce the concept of regularization, that permits to penalize large estimates that come from small sample sizes. The general idea is to add a penalty for large values of $b_{i}$ to the sum of squares equation that we minimize. So having many large $b_{i}$, make it harder to minimize. Regularization is a method used to reduce the effect of overfitting.


### IV. Regularized movie and user effect model

So estimates of $b_{i}$ and $b_{u}$ are caused by movies with very few ratings and in some users that only rated a very small number of movies. Hence this can strongly influence the prediction. The use of the regularization permits to penalize these aspects. We should find the value of lambda (that is a tuning parameter) that will minimize the RMSE. This shrinks the $b_{i}$ and $b_{u}$ in case of small number of ratings.


```{r lambdas, echo = TRUE}
lambdas <- seq(0, 10, 0.25)
rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  return(RMSE(predicted_ratings, validation$rating))
})
```


We plot RMSE vs lambdas to select the optimal lambda

```{r plot_lambdas, echo = TRUE}
qplot(lambdas, rmses)  
```

For the full model, the optimal lambda is:

```{r min_lambda, echo = TRUE}
  lambda <- lambdas[which.min(rmses)]
lambda
```

For the full model, the optimal lambda is: 5.25

The new results will be:


```{r rmse_results2, echo = TRUE}
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized movie and user effect model",  
                                     RMSE = min(rmses)))
rmse_results %>% knitr::kable()
```

\pagebreak

# Results

The RMSE values of all the represented models are the following:

```{r rmse_results3, echo = FALSE}
rmse_results %>% knitr::kable()
```

We therefore found the lowest value of RMSE that is 0.8648170.


# Discussion

So we can confirm that the final model for our project is the following:

$$Y_{u, i} = \mu + b_{i} + b_{u} + \epsilon_{u, i}$$

This model work well if the average user doesn't rate a particularly good/popular movie with a large positive $b_{i}$, by disliking a particular movie. 


# Conclusion

We can affirm to have built a machine learning algorithm to predict movie ratings with MovieLens dataset.
The regularized model including the effect of user is characterized by the lower RMSE value and is hence the optimal model to use for the present project.
The optimal model characterised by the lowest RMSE value (0.8648170) lower than the initial evaluation criteria (0.8775) given by the goal of the present project.
We could also affirm that improvements in the RMSE could be achieved by adding other effect (genre, year, age,..). Other different machine learning models could also improve the results further, but hardware limitations, as the RAM, are a constraint.

\pagebreak

# Appendix - Enviroment

```{r}
print("Operating System:")
version
```


